Division and Factorization of p-Adic Polynomials

### 2016-01-22 polynomials/muldivfact.tex
Adding some todos

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}
\DeclareMathOperator{\rank}{rank}

@@

\begin{rem}
\label{rem:unicity}
The divisor $A$ is uniquely determined by the conditions of 
Theorem~\ref{theo:slope-factor}. Indeed, consider two divisors
$A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
Then $L = \text{llcm}(A,A')$ is a divisor of $P$ as well and the 
slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
\NF(P)(d{-}1)$. Therefore $\deg L = d$ and $L$ differs from $A$ and
$A'$ by a multiplicative nonzero constant. Then, if $A$ and $A'$ 
share in addition the same leading coefficient, they must coincide.
\end{rem}

The rest of this subsection is devoted to the proof of the theorem.
If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
$P$ (resp. to the constant coefficient of $P$) and theorem is clear.
We then assume $1 < d < n$. Without loss of generality, we may further 
assume that the coefficient $a_d$ is equal to $1$. We set:
\begin{align*}
\lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d),
\end{align*}
so that $\kappa = \lambda_1 - \lambda_0$.
The existence of an extremal point of $\NP(P)$ located at abscissa
$d$ ensures that $\lambda_1 > \lambda_0$, \emph{i.e.} $\kappa > 0$.
For all indices $i$, we define:
$$\begin{array}{r@{\,\,}lr@{\,\,}l}
Q_i &= V_i P \div A_i, &
R_i & = V_i P \mod A_i = A_{i+1} - A_i, 
\smallskip \\
S_i &= P \mod A_i, &
T_i &= (1 - V_i B_i) \mod A_i
\end{array}$$
and when $\square$ is some letter, we put $\Delta \square_i = \square_{i+1} 
- \square_i$.
%We use the next Lemma to record some useful formulae.

@@

\begin{proof}
From $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1}$, we get
$- R_i B_{i+1}= \Delta B_i \cdot A_i + \Delta S_i$.
Hence, by consideration of degree, we obtain \eqref{eqdef:Biminus} 
and \eqref{eqdef:Siminus}.
On the other hand, from $V_i P = A_i Q_i + R_i=V_i (A_i B_i + S_i)$, we 
derive
\begin{equation}
\label{eq:ViBiQi} 
(V_i B_i -Q_i) \cdot A_i=R_i-V_i S_i. 
\end{equation} 
Thus $R_i=V_i S_i \mod A_i$.
Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
\mod A_i$, from what \eqref{eqdef:Si2} follows directly.
By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi} and
noting that $\deg R_i < \deg A_i = d$, we 
get \eqref{eqdef:Qiminus}.

@@

If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
is constant and the theorem is obvious. We then assume that $\lambda_0$
and $\lambda_1$ are both finite.
We notice that, when the polynomial is changed into $cP$ where $c$ is a 
nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
for $cP$. As a consequence we may assume that $P$ is normalized so that 
$\NF(P)(d) = d \lambda_0$. We define the function $\varphi : \R^+ \to \R 
\cup \{+\infty\}$ by:
\begin{equation}
\label{eqdef:functionphi}
\begin{array}{r@{\,\,}ll}
\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
\end{array}
\end{equation}
For a polynomial $Q \in K[X]$ of degree $n$, we further define:
\begin{align}
b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
\label{eqdef:bphi} \\
\text{and} \quad
b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
\quad \text{for } i \in \{0,1\}.
\label{eqdef:b0}
\end{align}
Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
With the normalization of $P$ we chose above, we have $b_0(P)
= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.

>>>>

\newcommand{\noopsort}[1]{}
\DeclareMathOperator{\NP}{NP}
\DeclareMathOperator{\HP}{HP}
\DeclareMathOperator{\PP}{PP}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\End}{End}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\val}{val}
\DeclareMathOperator{\pr}{pr}
\DeclareMathOperator{\tr}{Tr}
\DeclareMathOperator{\com}{Com}
\DeclareMathOperator{\Grass}{Grass}
\DeclareMathOperator{\Lat}{Lat}
\DeclareMathOperator{\round}{round}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\lcm}{lcm}

@@

\begin{rem}
\label{rem:unicity}
The divisor $A$ is uniquely determined by the conditions of 
Theorem~\ref{theo:slope-factor}. Indeed, consider two divisors
$A$ and $A'$ of $P$ such that $\NF(A) = \NF(A') = \NF(P)_{|[0,d]}$.
Then $L = \lcm(A,A')$ is a divisor of $P$ as well and the 
slopes of its Newton polygon are all at most $\lambda_0 = \NF(P)(d) - 
\NF(P)(d{-}1)$. Therefore $\deg L = d$ and $L$ differs from $A$ and
$A'$ by a multiplicative nonzero constant. Then, if $A$ and $A'$ 
share in addition the same leading coefficient, they must coincide.
\end{rem}

The rest of this subsection is devoted to the proof of the theorem.
If $d = n$ (resp. $d = 1$), the sequence $A_i$ is constant equal to
$P$ (resp. to the constant coefficient of $P$) and theorem is clear. \todo{Should this be $d=0$?}
We then assume $1 < d < n$. Without loss of generality, we may further 
assume that the coefficient $a_d$ is equal to $1$. We set:
\begin{align*}
\lambda_0 & = \NF(P)(d) - \NF(P)(d{-}1) \\
\lambda_1 & = \NF(P)(d{+}1) - \NF(P)(d),
\end{align*}
so that $\kappa = \lambda_1 - \lambda_0$.
The existence of an extremal point of $\NP(P)$ located at abscissa
$d$ ensures that $\lambda_1 > \lambda_0$, \emph{i.e.} $\kappa > 0$.
For all indices $i$, we define:
$$\begin{array}{r@{\,\,}lr@{\,\,}l}
Q_i &= V_i P \div A_i, &
R_i & = V_i P \mod A_i = A_{i+1} - A_i, 
\smallskip \\
S_i &= P \mod A_i, &
T_i &= (1 - V_i B_i) \mod A_i
\end{array}$$
and when $\square$ is some letter, we put $\Delta \square_i = \square_{i+1} 
- \square_i$.
%We use the next Lemma to record some useful formulae.

@@

\begin{proof}
From $P= A_i B_i + S_i= A_{i+1} B_{i+1} + S_{i+1}$, we get
$- R_i B_{i+1}= \Delta B_i \cdot A_i + \Delta S_i$.
Hence, by consideration of degree, we obtain \eqref{eqdef:Biminus} 
and \eqref{eqdef:Siminus}.
On the other hand, from $V_i P = A_i Q_i + R_i=V_i (A_i B_i + S_i)$, we 
derive
\begin{equation}
\label{eq:ViBiQi} 
(V_i B_i -Q_i) \cdot A_i=R_i-V_i S_i. 
\end{equation} 
Thus $R_i=V_i S_i \mod A_i$.
Hence $B_i R_i = (S_i - S_i T_i) \mod A_i$ and $S_i = B_i R_i + S_i T_i 
\mod A_i$, from which \eqref{eqdef:Si2} follows directly \todo{(there is a sign error here)}.
By definition of $V_i$, we get $\Delta V_i = V_i (1-V_i B_{i+1}) \mod A_{i+1}$ and consequently \eqref{eqdef:Viminus}.
We now write $1-Q_i= T_i + (V_i B_i - Q_i)$. Using \eqref{eq:ViBiQi} and
noting that $\deg R_i < \deg A_i = d$, we 
get \eqref{eqdef:Qiminus}.

@@

If $\lambda_0 = -\infty$ or $\lambda_1 = +\infty$, the sequence $(A_i)$
is constant and the theorem is obvious. We then assume that $\lambda_0$
and $\lambda_1$ are both finite.
We notice that, when the polynomial is changed into $cP$ where $c$ is a 
nonzero constant lying in a finite extension of $K$, the $A_i$'s are all 
multiplied by $c$ as well whereas the $B_i$'s and the $V_i$'s remained 
unchanged. Therefore, the theorem holds for $P$ if and only if it holds 
for $cP$. As a consequence we may assume that $P$ is normalized so that 
$\NF(P)(d) = d \lambda_0$. We define the function $\varphi : \R^+ \to \R 
\cup \{+\infty\}$ by:
\begin{equation} \label{eqdef:functionphi}
\varphi(x) = \begin{cases} \NF(P)(x) &\mbox{ if $x \leq d$} \\
\lambda_0 (x-d) + \NF(P)(d) &\mbox{ if $x > d$} \end{cases}
%\begin{array}{r@{\,\,}ll}
%\varphi(x) & = \NP(P)(x) & \text{for } x \leq d \smallskip \\
%& = \lambda_0 (x-d) + \NP(P)(d) & \text{for } x > d.
%\end{array}
\end{equation}
\todo{For $x>d$, why not say $\varphi(x) = \lambda_0 x$?} For a polynomial $Q \in K[X]$ of degree $n$, we further define:
\begin{align}
b_\varphi(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \varphi(x) 
\label{eqdef:bphi} \\
\text{and} \quad
b_i(Q) & = \min_{x \in [0,n]} \NF(Q)(x) - \lambda_i x
\quad \text{for } i \in \{0,1\}.
\label{eqdef:b0}
\end{align}
Set also $b_\varphi(0) = b_0(0) = b_1(0) = +\infty$ by convention.
With the normalization of $P$ we chose above, we have $b_0(P)
= b_\varphi(P) = 0$ and $b_\varphi(Q) \leq b_0(Q)$ for all polynomial
$Q$. Similarly $b_1(Q) \leq b_0(Q)$ for all $Q$.
